---
title: "Full Metabolomics Analysis Protocol"
author: "Abrar Al-Shaer"
date: "October 21, 2019"
output:
  html_document: default
  pdf_document: default
---

This file shows an example metabolomics full analysis protocol (given validated metabolomics data).
The example is using our lab's adipose tissue metabolomic data (from Nichole Reisdorph Lab). This was applied to the data for the manuscript by Pal et al. 2019 in the Shaikh Lab.


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown
Loading in pakcages
```{r cars}
rm(list=ls())
library("FactoMineR")
library(ggplot2)
library("factoextra")
library("corrplot")
library(ggfortify)
library(dplyr)
library(gplots)
library(ComplexHeatmap)
library(reshape2)
library(ggpubr)
```

Creating PCA Plots
```{r}
#for more detailed PCA code reference R script metabolites_PCA.R

#read in the file
#validated adipose data 10/15/19 - Con, HF, EPA
metData <- t(read.csv("Adipose Re-extraction Annotated_subset_rawData_Con_HF_EPA.csv", header = TRUE, fill = TRUE, row.names = 2))[c(2:12,14:16),] #selecting from this file the samples (rows) I wish to include in the analysis

#example of the file read in
metData[1:14,1:3]
```

```{r}
#convert the data frame to a numeric type
#2-77 are the columns that contain metabolite data
df2 <- data.frame(apply(metData[,2:77], 2, function(x) as.numeric(as.character(x))))
#assign the rownames to the dataframe
rownames(df2) <- rownames(metData)
#assign the column names to the dataframe
colnames(df2) <- colnames(metData)[2:77]
```

```{r}
#Defining active variables for PCA
metData.active <- df2[,c(1:length(df2))]
#running PCA
res.PCA <- PCA(metData.active, ncp = 5, graph=FALSE) #ncp = # of principal components to store from the PCA algorithm
```

Visualize Scree Plot
```{r}
#Scree plot
fviz_eig(res.PCA, addlabels = TRUE) 
```

Visualize metabolites that contribute to the dimensions in the PCA
```{r}
#provides a list of matrices containing all the results for the active variables 
#(coordinates, correlation between variables and axes, squared cosine and contributions)
var <- get_pca_var(res.PCA)
#Visualize correlation between the variables (columns) and principal componenet (PC)
#color by cos2 values: quality on the factor map
#select.var = select variables with top 30 contribution scores 
fviz_pca_var(res.PCA, col.var = "cos2", gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"), repel = TRUE, select.var = list(name = NULL, cos2 = NULL, contrib = 30)) #repel avoids overlapping labels
```

Visualize the Final PCA Graph
```{r}
#assign the PCA graphical output to ind.p
ind.p <- fviz_pca_ind(res.PCA, geom.ind = c("point", "text"), axes = c(1,2), labelsize = 6, pointshape = 20, pointsize = 5, repel = TRUE, select.ind = list(contrib = 50), col.ind = metData[,1], mean.point = FALSE, addEllipses = FALSE, legend.title = "Groups") 
#for visualization purposes you can select individual samples (select.ind) that contribute to the PCA the most (top 50, contrib = 50). This does not change was the calculations of the PCA
#axes = c(1,2) defines the principal components you want to visualize (i.e.PC1 & PC2)
#col.ind = column you would like to color the groups by (I color them by column 1 of my metData dataframe)

ggpubr::ggpar(ind.p, title = "PCA - Adipose Metabolites", xlab = "PC1 (59%)", ylab = "PC2 (17.8%)", legend.title = "Groups", legend.position = "top", ggtheme = theme_classic(), palette = "aaas")+
theme(plot.title=element_text(size=26))+theme(plot.subtitle=element_text(size=20))+
  theme(axis.title.y = element_text(size = 22))+theme(axis.title.x = element_text(size = 22))+
  theme(legend.title=element_text(size=25))+theme(legend.text=element_text(size=18))+font("xy.text", size = 22)
```

Testing for Statistical Significance & Calculating Fold Change
```{r}
#reading in transposed validated adipose datwith only HF & EPA groups
metData <- read.csv("Adipose Re-extraction Annotated_subset_rawData_HF_EPA_transpose.csv", header = TRUE, fill = TRUE, row.names = 1)[c(1:6,8:10),] #subset the samples (rows) you wish to include in the analysis (I'm excluding HF_105 because it's a clear outlier)

#assing a variables to the column names of the dataframe that contain METABOLITE names only (not other identifiers)
columns <- colnames(metData[,2:77])

#example of the file read in
metData[1:9,1:3] #only HF & EPA samples exist (i.e. 2 groups only) because we are also calculating fold change. So the dataframe must only contain 2 groups.
```

```{r}
#running the for loop that is iterating through all the columns & running the proper test & calculating FC
#make sure toreplace metData[1:5] with the rows for group 1 and metData[6:9] with the rows for group 2
datalist = list()
fold_change <- list()
i = 1
for(col in columns) #only loop through the column names specified above
{
  #print(col) #printing column name
  formula = as.formula(paste(col, "~ Diet")) #creating the comparison formula with the column
  
  if(shapiro.test(metData[,col])$p.value > 0.05){ 
    result = compare_means(formula,  data = metData, method = "t.test") #if data is normal
    if (result[[4]] < 0.05) { 
      datalist[[i]] = print(compare_means(formula,  data = metData, method = "t.test"))
      
      if(mean(metData[1:5,col]) > mean(metData[6:9,col])){ #if the mean of EPA is > than HF, then it's a (+)FC
        fold_change[[i]] = mean(metData[1:5,col])/mean(metData[6:9,col]) #mean of EPA / mean of HF 
      } else { fold_change[[i]] = (mean(metData[1:5,col])/mean(metData[6:9,col]))*-1 } #(-)FC -> EPA < than HF
    }
  } else { 
    result = compare_means(formula,  data = metData, method = "wilcox.test") #if the data is not normal
    if (result[[4]] < 0.05) { 
      datalist[[i]] = print(compare_means(formula,  data = metData, method = "wilcox.test"))
      
      if(mean(metData[1:5,col]) > mean(metData[6:9,col])){ #if the mean of EPA is > than HF, then it's a (+)FC
        fold_change[[i]] = mean(metData[1:5,col])/mean(metData[6:9,col]) #mean of EPA / mean of HF 
      } else { fold_change[[i]] = (mean(metData[1:5,col])/mean(metData[6:9,col]))*-1 } #(-)FC -> EPA < than HF
      
    } 
  }
  i = i+1
}

big_data1 = do.call(rbind, datalist) #combines all previous dataframes from for loop
big_data2 = data.frame(do.call(rbind, fold_change)) #contains fold change calculations
colnames(big_data2) = "Fold Change EPA:HF" #assign column name
big_data_final = dplyr::bind_cols(big_data1, big_data2) #combine all the dataframes together (stats & fold changes)
#write.csv(big_data_final, "Validated_Adipose_metabolites_FoldChange+signif_HF_EPA_unadjust.csv")

#adjust p-values
big_data_final$BH.adjust <- p.adjust(big_data_final$p, method = "BH")
#write.csv(big_data_final, "Validated_Adipose_metabolites_FoldChange+signif_HF_EPA_BH_adjust.csv")

#visualize big_data_final dataframe to show final output
big_data_final[1:5,1:10]
```

Generating Heatmaps
```{r}
#take the output file from the above code & add in a MetaboliteGroup column where you manually annotate the metabolite class of each metabolite (also add a Type column with an abbreviation of the class). Next take the fold change column & add in a Log2 FC column where you Log2 all the fold changes (for negative fold changes multiply the FC by -1 then calculate the Log2). 
metabolites <- read.csv("Validated_Adipose_metabolites_FoldChange+signif_HF_EPA_BH_adjust.csv", header = TRUE, fill = TRUE) 
#read in the transposed raw metabolomics data with any undesired samples removed (I removed HF_105 because it was an outlier)
metData <- read.csv("Adipose Re-extraction Annotated_subset_rawData_HF_EPA_transpose_rmHF105.csv", header = TRUE, fill = TRUE, row.names = 1)
```

Selecting only significant metabolites
```{r}
#extract metabolite names from the file that contains significant metabolites
compounds <- unique(as.character(metabolites$Metabolite))

#selects all occurances/matches for that list of significant compound
metData_select <- metData[,colnames(metData) %in% compounds] 
#add the sampleID label from the rownames of the previous dataframe (metData)
row.names(metData_select) <- rownames(metData) 

#transpose the metData_select dataframe
metData_select_t <- t(metData_select)

#write to a CSV the list of significant metabolites & their abundance values (raw data)
#write.csv(metData_select_t,"Validated_Adipose_metabolites_FoldChange+signif_HF_EPA_BH_adjust_metaboliteValues.csv")

#view output of dataframe
metData_select_t[1:6,1:9]
```

```{r}
#Take the above dataframe that you just wrote into a CSV and sort the dataframe & add the metabolite Group names (same metabolite group names as the metabolites dataframe). Also, filter the dataframe by fold changes above 1.5 or below -1.5
metabolites2 <- na.omit(read.csv("Validated_Adipose_metabolites_FoldChange+signif_HF_EPA_BH_adjust_metaboliteValues_1.5FC.csv", header = TRUE, fill = TRUE))
nams = metabolites2[,1] #names of the metabolites
#assign rownames to the dataframe based on metabolite names
rownames(metabolites2) = make.names(nams, unique=TRUE) #unique = TRUE to avoid duplicate names error

#view output of dataframe
metabolites2[1:5,1:15]
```

Creating the heatmap w/annotations
```{r}
# Annotation data frame
annot_df <- data.frame(Type = metabolites2$Type) #Log2FC = metabolites2$Log2FC (you can also add Log2FC annotation)
#Define colors for each levels of qualitative variables
#Define gradient color for continuous variables (FC)
col = list(Type = c("TG" = "lawngreen", "DG" = "darkgreen", "Amino Acids" = "orangered", "Ceramide" = "maroon1", 
                    "PE" = "yellow", "PA" = "blueviolet", "PC" = "goldenrod", "LysoPE" = "lemonchiffon4", 
                    "Macrolide" = "wheat1", "PS" = "turquoise1", "Galactolipid" = "rosybrown1", "Catechol" = "lavender")) #Log2FC = circlize::colorRamp2(c(-5,34), c("lightblue", "purple"))
# Create the heatmap annotation
ha <- HeatmapAnnotation(annot_df, col = col)

# Combine the heatmap and the annotation
Heatmap(scale(t(metabolites2[,7:15])), name = "Scaled \nAbundance", top_annotation = ha, row_names_gp = gpar(fontsize = 15), column_names_gp = gpar(fontsize = 13), cluster_columns = TRUE, cluster_rows = TRUE)

```

Fold Change Plot
```{r}
#subsetted (1.5 FC cutoff) validated adipose data
metData <- read.csv("Validated_Adipose_metabolites_FoldChange+signif_HF_EPA_BH_adjust_metaboliteValues_1.5FC.csv", header = TRUE, fill = TRUE)
#filter metData by only n3 PUFA containing metabolites
metData_sub <- metData[metData$n3PUFA_containing == "YES",]

#FOLD CHANGE BAR GRAPH
#color pallet
colors = c("TG" = "green3", "DG" = "darkgreen", "PC" = "goldenrod")

fc_plot <- ggbarplot(metData_sub, x = "Metabolite", y = "Log2FC",
          fill = "Type",           # change fill color by Type
          color = "white",            # Set bar border colors to white
          palette = colors,            # jco journal color palett is also an option. see ?ggpar
          sort.val = "none",          # If you want it sorted by FC then put (desc) to sort in descending order
          sort.by.groups = FALSE,     # Don't sort inside each group 
          x.text.angle = 90,          # Rotate vertically x axis texts
          ylab = "Log2FC EPA:HF",
          legend.title = "Type",
          rotate = TRUE,             #if Rotate = FALSE it will be a horizontal graph
          ggtheme = theme_minimal()
)

fc_plot + theme(axis.text.y = element_text(color = "black", size = 18),
                axis.text.x = element_text(color = "black", size = 18),
      axis.title.x = element_text(color = "black", face = "bold", size = 20),
      axis.title.y = element_text(color = "black", size = 18)) + 
  theme(legend.title = element_text(size=15, face="bold")) + theme(legend.text = element_text(size=15))
```

